Inferring generation-interval distributions from contact tracing data: \\ \emph{in prep, PRSB}

### 2019-05-06 doc/network.tex
trying to write methods

We refer to the second method as the individual-level method because it relies on individual contact information.
We model each infection $i$ from an infected individual $j$ as a non-homogeneous Poisson process between the time at which infector $j$ was infected ($t_j$) and the censorship time ($\tsub{t}{censor}$), where the time-varying Poisson rate is equal to the effective generation-interval distribution \citep{daley2007introduction}.
Then, the probability that an individual $j$ infects $n_j$ individuals between $t_j$ and $\tsub{t}{censor}$ is equal to
\begin{equation}
\frac{\RR^{n_j} \exp \big(- \RR G(\tsub{t}{censor} - t_j ;\theta) \big) \prod_{i=1}^{n_j} g(s_{i, j}; \theta)}{n_j!},
\end{equation}
where $s_{i,j}$ is the observed generation interval between infector $j$ and infectee $i$, and $\theta$ is a (vector) parameter of the generation-interval distribution $g$ (and the corresponding cumulative distribution function $G$).
We use gamma distribution to model the effective generation-interval distribution;
parameters $\theta$ correspond to the mean and shape parameters of a gamma distribution.

The entire data likelihood of the individual-level method can be written as:
\begin{equation}
\begin{aligned}
&\mathcal{L}(\RR, \theta\,|\, \mathbf{s}, \mathbf{t}, \mathbf{n}, \tsub{t}{censor})\\
&\quad=\prod_{j=1}^N \left(\RR^{n_j} \exp \big(- \RR G(\tsub{t}{censor} - t_j ;\theta) \big) \prod_{i=1}^{n_j} g(s_{i, j}; \theta) \right),
\end{aligned}
\end{equation}
where $N$ is the total number of infected individuals.
Here, we estimate parameters $\RR$ and $\theta$ by maximum likelihood.
\cite{forsberg2008likelihood} proposed a related approach based on discretized incidence reports.

\swp{Probabaly move simulations to appendix}
To test these methods, we simulate 100 epidemics with Ebola-like parameters on an empirical network and compare the estimates of reproductive number with empirical reproductive numbers, which we define as the average number of secondary cases generated by the first 100 infected individuals (\fref{cmp}).
\cite{trapman2016inferring} estimated the true reproductive number by dividing the total number of infected individuals in generations $2$ to $k+1$ by the total number of infected individuals in generations $1$ to $k$, where the reference generation $k$ is defined to be the first generation that contains at least 75 infected individuals;
however, individuals in later generations are potentially subject to stronger local depletion and even global depletion effects.
Therefore, we expect our empirical estimate to better reflect the infection process during the exponential growth phase.
\swp{TODO: Compare Trapman vs our estimate of empirical R0 in Appendix; Trapman estimate is much lower.}

@@

@@

>>>>

We refer to the second method as the individual-level method because it relies on individual contact information.
We model each infection $i$ from an infected individual $j$ as a non-homogeneous Poisson process between the time at which infector $j$ was infected ($t_j$) and the censorship time ($\tsub{t}{censor}$), where the time-varying Poisson rate at time $t$ is equal to $\RR g(t - t_j)$, where $g(t)$ is the effective generation-interval distribution \citep{daley2007introduction}.
We use a gamma distribution (parameterized by its mean and shape) to model the effective generation-interval distribution.
Then, the probability that an individual $j$ infects $n_j$ individuals between $t_j$ and $\tsub{t}{censor}$ is equal to
\begin{equation}
\frac{\RR^{n_j} \exp \big(- \RR G(\tsub{t}{censor} - t_j ;\theta) \big) \prod_{i=1}^{n_j} g(s_{i, j}; \theta)}{n_j!},
\end{equation}
where $s_{i,j}$ is the observed generation interval between infector $j$ and infectee $i$, and $\theta$ is a (vector) parameter of the generation-interval distribution $g$ (and the corresponding cumulative distribution function $G$).

The entire data likelihood of the individual-level method can be written as:
\begin{equation}
\begin{aligned}
&\mathcal{L}(\RR, \theta\,|\, \mathbf{s}, \mathbf{t}, \mathbf{n}, \tsub{t}{censor})\\
&\quad=\prod_{j=1}^N \left(\RR^{n_j} \exp \big(- \RR G(\tsub{t}{censor} - t_j ;\theta) \big) \prod_{i=1}^{n_j} g(s_{i, j}; \theta) \right),
\end{aligned}
\end{equation}
where $N$ is the total number of infected individuals.
Here, we estimate parameters $\RR$ and $\theta$ by maximum likelihood.
Since the estimate of $\RR$ is sensitive to under-reporting, we use the estimated generation-interval distribution to infer the effective reproductive number from the estimated growth rate.
\cite{forsberg2008likelihood} proposed a related approach based on discretized incidence reports.

To test these methods, we simulate 100 epidemics with Ebola-like parameters on an empirical network \citep{leskovec2016snap} and compare the estimates of reproductive number with empirical reproductive numbers, which we define as the average number of secondary cases generated by the first 100 infected individuals (\fref{cmp}).
\cite{trapman2016inferring} estimated the true reproductive number by dividing the total number of infected individuals in generations $2$ to $k+1$ by the total number of infected individuals in generations $1$ to $k$, where the reference generation $k$ is defined to be the first generation that contains at least 75 infected individuals;
however, individuals in later generations are potentially subject to stronger local depletion and even global depletion effects.
Therefore, we expect our empirical estimate to better reflect the infection process during the exponential growth phase.
\swp{TODO: Compare Trapman vs our estimate of empirical R0 in Appendix; Trapman estimate is much lower.}

@@

The generation-interval-based approach to estimating the reproductive number assumes that an epidemic grows exponentially \citep{wallinga2007generation, park2019practical}.
In practice, heterogeneity in population structure can lead to subexponential growth \citep{chowell2016characterizing, viboud2016generalized}.

@@

\subsection{Deterministic SEIR model}

\swp{CITE}
To study the effects on right-censoring on the observed generation intervals, we use the deterministic Susceptible-Exposed-Infectious-Recovered (SEIR) model.
The SEIR model describes how a disease spreads in a homogeneously mixing population; it assumes that infeceted individuals become infectious after a latent period.
While we use the standard SEIR model, which assumes exponentially distributed latent and infectious period, throughout this paper, we present the extended version of the model, which allows for Erlang-distributed latent and infectious periods (we present some results based on the extended model in Appendix): 
\begin{equation}
\begin{aligned}
\frac{dS}{dt} &= - \beta S \frac{\sum_{k=1}^{n_I} I_k}{N}\\
\frac{dE_1}{dt} &= \beta S \frac{\sum_{k=1}^{n_I} I_k}{N} - n_E \sigma E_1\\
\frac{dE_m}{dt} &= n_E \sigma (E_{m-1} - E_m) && \text{for } m = 2, 3, \dots, n_E\\
\frac{dI_1}{dt} &= n_E \sigma E_m - n_I \gamma I_1\\
\frac{dI_n}{dt} &= n_I \gamma (I_{n-1} I_n) && \text{for } n = 2, 3, \dots, n_I\\
\end{aligned}
\end{equation}
where $S$ is the number of susceptible individuals, $E_m$ is the number of exposed individuals in the $m$-th compartment, $I_n$ is the number of infectious individuals in the $n$-th compartment, and $R$ is the number of recovered individuals.
Parameters of the model are specified as follows: $N$ is the total population size, $\beta$ is the transmission rate, $1/\sigma$ is the mean latent period, and $1/\gamma$ is the mean infectious period.

\subsection{Stochastic SEIR model}

\swp{What do I cite here? Gillespie?}
We use individual-based simulations of the stochastic SEIR model on a network to study variation in realized generation intervals.
Every node of a network is numbered from 1 to $N$, where $N$ is the total number of nodes. 
Each node represents an individual.
Likewise, an edge of a network represents acquaintanceship; infected individuals can only contact their acquaintances.

First, we begin by randomly selecting initially infected individuals.
For each infected individual $i$, we randomly draw a random number $E_i$ from an Erlang distribution with mean $1/\sigma$ and shape $n_E$.
Given a total contact rate of $k_i \lambda$, where $k_i$ is the number of acquaintances and $\lambda$ is the per-pair contact rate, and a transition rate of $n_I \gamma$ between each infectious stages, 
the total number of contacts that an indiviudal $i$ in infectious stage $j$ makes before moving to infectious stage $j+1$ follows a geometric distribution with probability $n_I\gamma/(k_i \lambda + n_I\gamma)$:
\begin{equation}
C_{i, j} \sim \mathrm{Geometric}\left(\frac{n_I\gamma}{k_i \lambda + n_I\gamma}\right).
\end{equation}
For each contact, a contactee is uniformly sampled from the set of acquaintances of the individual $i$.

\subsection{Empirical network}
